Terrain is a stronger predictor of peat depth than airborne radiometrics in Norwegian landscapes

### 2024-12-05 ms/ms.Rmd
ms: revise abstract

  Peatlands are Earth's most carbon-dense terrestrial ecosystems and crucial stores of soil carbon. 
  Their carbon density results largely from their depth, which can exceed ten meters. 
  However, peat depth is poorly mapped at scales relevant for land management and carbon accounting. 
  We evaluated whether digital soil mapping using remotely sensed data could improve existing maps of peat depth in Norway. 

@@

  We measured peat depth at 372 and 1878 locations using manual probing and ground-penetrating radar, then trained Random Forest models to predict depth at 10 m resolution from various combinations of predictors. 

@@

  Terrain variables were more important predictors than radiometric variables, with elevation and valley bottom flatness showing the strongest relationships to depth. 
  Adding radiometric predictors improved model performance at one site but degraded it at the other. 
  Our models improved upon the national peat depth map, but the weak relationships between peat depth and remotely sensed variables demonstrate limitations to mapping without extensive field measurements. 
  Models trained only on known peatlands could not reliably identify peat presence elsewhere, highlighting the need to map peat occurrence and depth together.

>>>>

  Peatlands are Earth's most carbon-dense terrestrial ecosystems and their carbon density varies with the depth of the peat layer. 
  Accurate mapping of peat depth is crucial for carbon accounting and land management, yet existing maps lack the resolution and accuracy needed for these applications. 
  This study evaluates whether digital soil mapping using remotely sensed data can improve existing maps of peat depth in Norway. 

@@

  We measured peat depth at 372 and 1878 locations and trained Random Forest models to predict depth at 10 m resolution. 

@@

  Terrain variables were better predictors than radiometric variables, with elevation and valley bottom flatness showing the strongest relationships to depth. 
  Radiometric data showed inconsistent value, improving predictions at one site while degrading performance at the other. 
  Our models improved upon the national peat depth map, but weak relationships with remotely sensed variables make peat depth hard to predict. 
  These findings suggest that while digital soil mapping can enhance existing peat depth maps, reliable carbon stock assessments still require strategic field sampling. 
  Furthermore, the inability of models to identify peat presence outside known peatlands highlights the need for integrated mapping of peat occurrence and depth.

### 2024-12-05 ms/ms.tex
ms: revise abstract

\begin{abstract}
Peatlands are Earth's most carbon-dense terrestrial ecosystems and crucial stores of soil carbon.
Their carbon density results largely from their depth, which can exceed ten meters.
However, peat depth is poorly mapped at scales relevant for land management and carbon accounting.
We evaluated whether digital soil mapping using remotely sensed data could improve existing maps of peat depth in Norway.
Specifically, we assessed the predictive value of LiDAR-derived terrain variables and airborne radiometric data at two contrasting sites.
We measured peat depth at 372 and 1878 locations using manual probing and ground-penetrating radar, then trained Random Forest models to predict depth at 10 m resolution from various combinations of predictors.
The best models achieved mean absolute errors of 60 and 56 cm, explaining one-third of the variation in peat depth.
Terrain variables were more important predictors than radiometric variables, with elevation and valley bottom flatness showing the strongest relationships to depth.
Adding radiometric predictors improved model performance at one site but degraded it at the other.
Our models improved upon the national peat depth map, but the weak relationships between peat depth and remotely sensed variables demonstrate limitations to mapping without extensive field measurements.
Models trained only on known peatlands could not reliably identify peat presence elsewhere, highlighting the need to map peat occurrence and depth together.
\end{abstract}

>>>>

\begin{abstract}
Peatlands are Earth's most carbon-dense terrestrial ecosystems and their carbon density varies with the depth of the peat layer.
Accurate mapping of peat depth is crucial for carbon accounting and land management, yet existing maps lack the resolution and accuracy needed for these applications.
This study evaluates whether digital soil mapping using remotely sensed data can improve existing maps of peat depth in Norway.
Specifically, we assessed the predictive value of LiDAR-derived terrain variables and airborne radiometric data at two contrasting sites.
We measured peat depth at 372 and 1878 locations and trained Random Forest models to predict depth at 10 m resolution.
The best models achieved mean absolute errors of 60 and 56 cm, explaining one-third of the variation in peat depth.
Terrain variables were better predictors than radiometric variables, with elevation and valley bottom flatness showing the strongest relationships to depth.
Radiometric data showed inconsistent value, improving predictions at one site while degrading performance at the other.
Our models improved upon the national peat depth map, but weak relationships with remotely sensed variables make peat depth hard to predict.
These findings suggest that while digital soil mapping can enhance existing peat depth maps, reliable carbon stock assessments still require strategic field sampling.
Furthermore, the inability of models to identify peat presence outside known peatlands highlights the need for integrated mapping of peat occurrence and depth.
\end{abstract}
